---
title: "SMPA - Grouping Assignment"
author: "YourNameHere"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r}

#load libraries
library(tidyverse)
library(lubridate)

#load data
payments <- readRDS("data/trump_campaign_payments.rds")

```

## Introduction

**What data will you be working with?**

A dataset of disbursements (payments and purchases) by the Donald J. Trump for President campaign committee during the 2020 election cycle. The included payments are from 2017 and 2018.

This data and examples were developed as part of the NYT Open training, which we've adapted from here. Note that the dataset has been alterted for instructional purposes, so doesn't represent the original FEC data in its entirety.

In the environment tab in the upper right, click on the `payments` dataset to open the RStudio viewer to examine the data. You can also run the chunk below as well to bring up a slice of the data.

```{r}

head(payments)


```

## Before you begin

Take a moment to look at the various categories included for each disbursement.

-   Notice that there are several date-related fields:

    -   FEC election cycle

    -   report year

    -   election type

    -   disbursement date

-   There are a few fields about the recipient. Along with city and state, you have:

    -   the recipient name

    -   the payee's first and last name (an individual who was paid for the "recipient") and

    -   the entity type (individual v. organization)

## Initial Narrative Questions

Without doing any analysis yet, what are three questions you might try to answer using this dataset? Write your narrative answers in the spaces below.

1.  What disbursement did the Trump campaign spend the most money on?

2.  How much money did they spend? How about in total?

3.  How did the type of disbursement change over time?

## Coding Assignment Questions

**Please write your R tidyverse code to help answer the questions below in the provided R chunks. You can make more than one chunk if you prefer.**

**Also, describe what you're doing either in text out "in the open" above the chunk, or in comments (prefaced by #) in the R chunk itself.**

**You'll need to use grouping, filtering, or a combination of both to find the answers. Remember that R is case-sensitive when you refer to any text values within the columns**

<br>

How much was spent in the 2017 vs. 2018 reporting year?

I start with payments, then group by the year, then summarize with the sum of the disbursement amount to get the total.

```{r}

payments %>% 
  group_by(report_year) %>% 
  summarize(sum(disbursement_amount))





```

Tally the total overall disbursement amounts for each month (distinguishing between say, Oct. 2018 v. Oct. 2019). In which month was spending the highest?

group by both disbursement_month and disbursemeant year, then get the sum for the amount. It looks like January 2017 was the highest.
```{r}

payments %>% 
  group_by(disbursement_month, disbursement_year) %>% 
  summarize(sum(disbursement_amount))




```

Which recipient (recipient_name) was paid the most in the 2020 election cycle?

Group by receipt name and then summarize the sum of disbursement amount. ABATE MEDIA LLC was paid the most in the 2020 election cylcle.
```{r}


payments %>% 
  group_by(recipient_name) %>% 
  summarize(sum(disbursement_amount))




```

Look at what this recipient was paid for. What service or good (see the "disbursement_description") were they paid the most for?

Filter for that company first, then group. AUDIO VISUAL SERVICES is the what they paid the most for.
```{r}

payments %>% 
  filter(recipient_name == "ABATE MEDIA LLC") %>% 
  group_by(disbursement_description) %>% 
  summarize(sum(disbursement_amount))





```

What's the most common category overall (see "disbursement_description") the campaign spent money on?

"MERCHANT FEES" is the most that the campaign spent money on, which is very weird.
```{r}

payments %>% 
  group_by(disbursement_description) %>% 
  summarize(n = n())

payments %>% 
  group_by(disbursement_description) %>% 
  summarize(n = n()) %>% 
  arrange(desc(n))

```

Who were the 5 biggest recipients of campaign spending on "LEGAL CONSULTING"..?

Filter, group, summarize by sum, arrange. The 5 biggest are 

JONES DAY	4390351.66	
LAROCCA HORNIK ROSEN GREENBERG & BLAHA LLP	404774.48	
KASOWITZ, BENSON, TORRESS & FRIEDMAN	322614.29	
THE LAW OFFICES OF ALAN S. FUTERFAS	287924.00	
MCDERMOTT WILL & EMERY
```{r}

payments %>% 
  filter(disbursement_description == "LEGAL CONSULTING") %>% 
  group_by(recipient_name) %>% 
  summarize(n = sum(disbursement_amount)) %>% 
  arrange(desc(n))





```


Who are the top 5 individuals (vs. organization) were paid by the campaign committee?

group, summarize, arrange. The top 5 payees are:

GLASSNER	511280.00			
DOLLMAN	389541.09			
CENTINELLO	214934.25			
CARR	160000.00			
PENCE	104244.03
```{r}


payments %>% 
  group_by(payee_last_name) %>% 
  summarize(n = sum(disbursement_amount)) %>% 
  arrange(desc(n))





```


For the indvidual who was paid the most, which payment category (disbursement_description) was the top one for him or her? And how much money were they paid in that category?

The largest expense Glassner was paid was expense type PAYROLL, and they got paid, 499000.00 for that type of expense.
```{r}

payments %>% 
  filter(payee_last_name == "GLASSNER") %>% 
  group_by(disbursement_description) %>% 
  summarize(n = n()) %>% 
  arrange(desc(n))

payments %>% 
  filter(payee_last_name == "GLASSNER") %>% 
  group_by(disbursement_description) %>% 
  summarize(n = sum(disbursement_amount)) %>% 
  arrange(desc(n))






```
